Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-07-17 ms/manuscript.Rnw
Changes to figures and manuscript

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling. Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth. We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals encompassed within 203 species. We show that both growth-independent and growth-dependent hazards are significant contribtors to tree mortality at Barro Colorado Island. Moreover, we show that species traits -- wood density, maximum dbh and shade intolerance -- all variously affect rates of tree mortality but ultimately only account for a small fraction of the total species variability.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive a instantaneous mortality function for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop a mortality function that not only integrates individual, species, and temporal effects, but also partitions the contributions of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations. Moreover, we show that while these three traits can affect both growth-dependent and growth-independent hazards, they explain only only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}. This variation has considerable consequences for forest structure and dynamics. For example, the death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun. In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}. Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change.

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}. This requirement arises in part because the observed outcome -- alive vs. dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also known as the ``hazard function''. In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ can be thought of as the cumulative hazard an individual obtains between $t_1$ and $t_2$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale.
\begin{equation}
S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
\end{equation}
As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i$) from binary data, large sample sizes collected over long durations are required. Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.

Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality. Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}. However, these influences have not yet been integrated into a common hazard function \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.

@@

A specific challenge in developing hazard functions for plants is to estimate the independent contributions of growth-dependent and growth-independent hazards on an individual's overall hazard. While plants die via many causes, these broadly fall into two main types of hazards. The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as windfall or fire. An individual's total hazard is therefore the sum of these independent hazards.

A further challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth. Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies broadly support this theory, with many indicating exponential-like declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}. By contrast, across species, there is a strong trade-off between growth and survival, with faster-growing species exhibiting higher mortality rates relative to slower-growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via species traits with antagonistic effects, such as wood density. Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

Here, we attempt to reconcile intra- and inter-specific variation as well as account for growth-dependent and growth-independent hazards. We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Growth independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits (wood density, $\rho$, shade-intolerance, $\upsilon$, and maximum dbh, $\psi$), as well as a species random effect that captures any remaining species-level differences that is not otherwise accounted for by these traits (for details see Methods).

To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across seven iterative stages of model development.

In stage 1, we ask whether mortality rates vary substantially among censuses, and thus, whether a census effect is required. We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e. a null model) and compared it to the same model with a census random effect.
In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these. Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig. \ref{fig1}B) best predicts the survival data. The simplest form (Fig. \ref{fig1}B, left) assumes a constant growth-independent hazard rate. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases. The third form (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses previously proposed (e.g. \cite{Kobe:1995tw, Wyckoff:2000un}), including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared. We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
In stage 3, we examine whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards. Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density. This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density. Using the final ``best'' model we also conducted post-hoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A). Individual census effects can be found in Table S1.

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
A) Five sequential stages of model selection with each stage increasing in complexity. Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Predictive skill for wood density parameter combinations. Shading represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. For both panels symbols represent functional form: null (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species growth-independent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure}

\subsection*{Wood density and other species effects}
Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig. \ref{fig4} A,B), and thus higher probability of death across a 1-yr period (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species. For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$. Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig. \ref{fig2}A). See Figs. S3-S5 for species-level parameter estimates.

\begin{figure}[t]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in instantaneous mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing growth rate. A,C) Estimated curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.}
\label{fig4}
\end{figure}

\subsection*{Parameter contribution to predicted variation}
To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Fig. \ref{fig5}). We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model. Species effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality. Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation. Census accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation. The growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig. \ref{fig5}).

\subsection*{Mortality rates, maximum species size and species light requirements}
To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits. We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement; Fig. \ref{fig6} A,C,E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade. Correlations between estimated species parameters and their associated maximum DBH were weak (Fig. \ref{fig6}).

\begin{figure}[t]
\includegraphics{figures/fig5.pdf}
\caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects. Effects are not mutually exclusive thus sum to more than 1.0. Note "wood density" is a subset of the overall "species" effect.}  
\label{fig5}
\end{figure}

@@

Additionally, the partitioning of mortality into growth-dependent and growth-independent effects allowed us to estimate the proportion of variation attributed to each. Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig. \ref{fig5}). This suggests deficiencies in carbon budget are a major contributor to tree death on BCI.

@@

While wood density effects are now being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig. \ref{fig4}-\ref{fig5}). This means that such models are likely to severely underestimate the true variability in mortality rates. Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.

\begin{figure}[t]
\includegraphics{figures/fig6.pdf}
\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (left) and species maximum DBH (right). Points are means ($\pm$ 95\% credible intervals) for each of the 203 species used in this study. Blue line shows the average trendline as determined from standard linear regression. Gap index is a species level index of the light required for recruitment. Low gap index indicates the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
\label{fig6}
\end{figure}

>>>>

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling. Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also, for the first time, simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth. We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals encompassed within 203 species. We show that both growth-independent and growth-dependent hazards are significant contribtors to tree mortality at Barro Colorado Island. Moreover, we show that species traits -- wood density, maximum dbh and shade intolerance -- all variously affect rates of tree mortality but ultimately only account for a small fraction of the total species variability.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related and growth-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations. Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}. This variation has considerable consequences for forest structure and dynamics. For example, the death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun. In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}. Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of increased climate-driven forest die-off \cite{Anderegg:2012ep}.

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}. This requirement arises in part because the observed outcome -- alive vs. dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality rate for individual $i$ at time $t$, also known as the ``hazard function or hazard rate''. In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ can be thought of as the cumulative hazard an individual obtains between $t_1$ and $t_2$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale.
\begin{equation}
S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
\end{equation}

As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required. Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.

Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality. Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits such as wood density (REFS), maximum size (REFS) and measures of shade tolerance (REFS) \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}. However, these intra- and inter-specific effects have not yet been integrated into a common hazard function that accounts for the two fundamental hazards faced by plants -- death due to low carbon budget and death due to stochastic events that are unrelated to carbon budget (e.g. fire, windthrow) \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.

@@

A specific challenge in developing biologically plausible hazard functions for plants is the need to estimate the independent contributions of carbon-dependent and carbon-independent hazards on an individual's overall hazard. While plants die via many causes, these broadly fall into two main hazards. The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are carbon-independent hazards, where plants die because of stochastic events that are unrelated to their growth rate, such as windfall or fire. An individual's total hazard at a given time point is therefore the sum of these two independent hazards, and is contrary to many traditional survival analyses done for plants which assume these hazards are proportional to eachother (REF).

A further challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth. Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies broadly support this theory, with many indicating exponential-like declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}. By contrast, across species, there is a strong trade-off between growth and survival, with faster-growing species exhibiting higher mortality rates relative to slower-growing species \cite{Poorter:2008iu, Wright:2010fl}. This trade-off between growth and survival may arise via species traits with antagonistic effects, such as wood density. Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

Here, we attempt to develop a empirical model capable of reconciling these conflicting growth-mortality relationships observed among individuals and between species while also explicitly estimating the independent contributions of carbon-dependent and independent hazards. We achieve this by evaluating the following, unified hazard function that incorporates individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Growth independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for the independent effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), shade-intolerance ($\upsilon$) and maximum dbh ($\psi$). We also include species random effects that captures any remaining species-level differences that are not otherwise accounted for by species traits (for details see Methods).

To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_training_full)} individuals from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across seven iterative stages of model development.

In stage 1, we ask whether mortality rates vary substantially among censuses, and thus, whether a census effect is required. We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e. the null model) and compared it to the same model but with a census random effect.
In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these. Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig. \ref{fig1}B) best predicts the survival data. The simplest form (Fig. \ref{fig1}B, left) assumes a constant growth-independent hazard rate. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases. The third form (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses previously proposed (e.g. \cite{Kobe:1995tw, Wyckoff:2000un}), including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared. We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
In stage 3, we examine whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards. Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density. This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density. Using the final ``best'' model we also conducted post-hoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A). Individual census effects can be found in Table S1.

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
A) Five sequential stages of model selection with each stage increasing in complexity. Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Predictive skill for wood density parameter combinations. Shading represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. For both panels symbols represent functional form: null (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure*}

@@

\begin{figure*}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species growth-independent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure*}

\subsection*{Species traits}
Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig. \ref{fig4} A,B), and thus higher probability of death across a 1-yr period (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species. For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$. Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig. \ref{fig2}A). See Figs. S3-S5 for species-level parameter estimates.

To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits. We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade. Correlations between estimated species parameters and their associated maximum DBH were weak).

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing growth rate. A,C) Estimated curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.}
\label{fig4}
\end{figure}

\subsection*{Parameter contribution to predicted variation}
To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Table XX). We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model. Species effects contributed...

% \begin{figure}[t]
% \includegraphics{figures/fig5.pdf}
% \caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects. Effects are not mutually exclusive thus sum to more than 1.0. Note "wood density" is a subset of the overall "species" effect.}  
% \label{fig5}
% \end{figure}

@@

Additionally, the partitioning of mortality into growth-dependent and growth-independent effects allowed us to estimate the proportion of variation attributed to each. Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that the growth-dependent hazard accounted for XX of the total predicted variability in mortality rates (Table X). This suggests deficiencies in carbon budget are a major contributor to tree death on BCI.

@@

While wood density effects are now being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig. \ref{fig4}, Table X). This means that such models are likely to severely underestimate the true variability in mortality rates. Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.

We found that wood density only explained XX of the total XX variation explained by species effects, suggesting that other traits are also affecting tree mortality rates. Post-hoc analyses suggest this unexplained variation was in part related to a species' light demand, but not its maximum height. Both traits have been proposed major axes of inter-specific variation in tropical rainforests \cite{Poorter:2006tq} (Fig. XX). Specifically, light demanding species (i.e. high gap index) had higher growth-independent hazard rates and were more susceptible to dying as a result of low growth, relative to those that readily recruit in shade (i.e. low gap index), supporting past findings \cite{Ruger:2011cv}. By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting \cite{Poorter:2006tq} and supporting \cite{Wright:2010fl} previous results. Future research should therefore resolve how other traits influence growth-dependent and growth-independent hazards, and identify the combination of traits needed to improve model predictive skill.
